Learning CNN filters from user-drawn image makers for coconut-tree image classification

### 2020-06-12 main.tex
Updates from Overleaf

\section{Introduction}
Deep learning has proven to be applicable to different tasks, from image classification to data synthesis \cite{goodfellow2016deep}. In remote sensing, the applications may involve segmentation of terrain images \cite{kemker2018algorithms, kampffmeyer2016semantic, hamaguchi2018effective}, building identification \cite{xu2018building, lu2018detecting, liu2018multilevel}, and deforestation monitoring \cite{bragilevsky2017deep}, for instance. In this work, we are interested in 
 identifying species of trees from aerial images. The topic is important for land-use classification, plantation monitoring, and damage assessment of natural disasters. As the plantations can span very large areas, the manual identification of each tree is costly, tedious, and error-prone, and so automatic classification methods are necessary.

Classification of tree species in aerial images has been actively investigated~\cite{fassnacht2016review}. In \cite{puttemans2018comparing} and~\cite{aparna2018cnn}, the authors present automatic solutions to detect coconut trees based on convolutional neural network (CNN) models. Despite these recent advances, CNN models usually require considerable human effort in image annotation to create large training sets. Vargas-Muñoz et al.~\cite{8899005} propose an active learning approach to mitigate the problem. The method explores data projection techniques to allow simultaneous annotation of multiple patches as having or not one coconut tree. It then uses a CNN to identify the candidate patches with one coconut tree. We adopt another alternative -- the design of simplified CNN models from small training sets with user interaction, being the user an expert in the design of CNN models.

@@

In order to reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called FLIM (\emph{Feature Learning from Image Markers}), that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, a few training images, and draws markers in regions of the input images that best discriminate the classes. We explain how the classes may appear in distinct clusters of a local feature space defined by patches as extracted from marker pixels at the input of each convolutional layer. FLIM defines the weights of the filters as the centroids of the clusters. Such filters are expected to activate discriminant regions in the output of the given convolutional layer, and the process repeats in a layer-by-layer fashion. The architecture of the CNN could also be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}.

We demonstrate the advantages of this interactive technique on binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of  marker selection.  Second, very few images per class (e.g., less than five) seem to be  enough. Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers.  Fourth, FLIM is application-independent, and so it might be useful in other image classification problems.

@@

A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation by eliminating negative activations, and aggregate the resulting activations (pooling and flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success. Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and use fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually rely on  weight optimization by backpropagation, which requires a higher number of annotated images as deeper is the model. In FLIM, we eliminate the need of backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images $I$ to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect\left[P_I(p)\right] \cdot  \vect\left[F\right]$, where $P_I(p)$ is a patch with $k\times k$ pixels and $m$ image channels around $p$, $F$ is a filter with $k\times k\times m$ weights, and $\vect\left[.\right]$ is the vectorization operation. A filter $F$ can discriminate a class $i\in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect\left[P_I(p)\right]$ of class $i$ and negative results for patterns from other classes. That is, $\vect\left[F\right]$ is the normal vector of a hyperplane in $\Re^{k\times k\times m}$ which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect\left[P_I(p)\right]$ from all $c$ classes.  
\begin{eqnarray}
{\cal P} & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p)\in [1,c]} P_I(p)
\label{eq.pset}
\end{eqnarray}
The patch set ${\cal P}$ defines samples from all classes in $\Re^{k\times k\times m}$ and a clustering operation on ${\cal P}$ creates groups with similar local patterns. Ideally, each group should contain patterns of a same class. However, some classes might share similar patterns. We should at least expect that their patterns be detected in different image regions. The groups must be shifted to the origin of the feature space by subtracting the mean vector of the samples in ${\cal P}$. Optionally, the standard deviation of each feature can be computed and used for feature standardization. We call this operation \emph{patch-based normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$ and we force norm $\|\vect\left[F\right]\|=1$ to avoid preferences among them. The centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane with normal $\vect\left[F\right]$. Figure~\ref{fig:filter}a illustrates three groups with color-coded samples from two classes in a hypothetical 2D feature space. The path-based normalization followed by convolution with filters obtained as centroids of the three groups creates a new 3D feature space (Figure~\ref{fig:filter}b), in which the classes can be more easily detected by two filters.

\begin{figure}[!ht]
  \centering
  \subfloat[\label{fig:ex-groups}]
  {\includegraphics[width=.35\linewidth]{figures/example_groups/groups_centered}}
  ~
  \subfloat[\label{fig:ex-groups-after}]{\includegraphics[width=.45\linewidth]{figures/example_groups/groups3d_centered}
  }
  \caption{(a) A 2D feature space with color-coded samples from two classes distributed in three groups. (b) After patch-based normalization and convolution with filters obtained from each group, the classes can be more easily detected by two filters in a new 3D feature space.}
  \label{fig:filter}
\end{figure}

Each convolutional layer is trained individually, one layer at a time, to find its filter set ${\cal F}$. The number of filters per layer depends on the clustering technique. We use $K$-means, then the user must specify the number $K=|{\cal F}|$ of filters. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max-pooling operation to aggregate local information. Note that, apart from the initial marker selection, the training process is automatic. We have preserved the convolutional layers' output dimensions, such that we can use the same markers used in the input layer to find the filters of layer $L+1$ from the resulting images obtained at the output of layer $L$. Figure~\ref{fig:tree} shows one example of a simple network projected by FLIM for the experiments of the next section.

@@

  \begin{tikzpicture}
    [
      grow                    = down,
      sibling distance        = 5em,
      level distance          = 3em,
      edge from parent/.style = {draw=black},
      every node/.style       = {treenode},
      sloped
    ]
    \node [treenode, fill=tealDeer, label = {[label distance = 0.5em]}] (v0) {Conv$(60 \times 7 \times 7)$}
      child {
        child {
          child {
                child {
                  node [treenode, fill=navyPurple] {Classifier}
                  edge from parent
                }
                node [treenode, fill=tealDeer] {BatchNorm}
                edge from parent
           }
          node [treenode, fill=tealDeer] {MaxPool$(3\times3)$}
          edge from parent
        }
        node [treenode, fill=tealDeer, label = {[label distance = 0.5em] }] {ReLU} 
        edge from parent
      };

@@

\section{Experiments}
We use a dataset formed from patches of aerial images made by \cite{8899005} from imagery provided by WeRobotics. The imagery with a resolution of 8cm was captured by the World Bank's UAVs for Disaster Resilience Program, collaborating with WeRobotics and OpenAerialMap in the Kingdom of Tonga in October 2017. The Humanitarian OpenStreetMap community provided the annotation. The dataset consists of 13587 patches of dimensions $90 \times 90$ of coconut (10268) and non-coconut (3319). The latter class has a large variability. The problem of interest is to classify the images in one of the two classes. We only report experiments with one dataset due to the space limitation.

@@

      FLIM & $0.863 \pm 0.002$ & $0.849 \pm 0.005$ & $0.854 \pm 0.004$\\
    FLIM-FT &  &  & \\

    VGG &  &  & \\
    VGG-FT & $0.872 \pm 0.007$ & $0.844 \pm 0.015$ & $0.851 \pm 0.014 $ \\

>>>>

\section{Introduction}
Deep learning has proven to be applicable to different tasks, from image classification to data synthesis \cite{goodfellow2016deep}. In remote sensing, the applications may involve segmentation of terrain images \cite{kemker2018algorithms, kampffmeyer2016semantic, hamaguchi2018effective}, building identification \cite{xu2018building, lu2018detecting, liu2018multilevel}, and deforestation monitoring \cite{bragilevsky2017deep}, for instance. In this work, we are interested in identifying species of trees from aerial images. The topic is important for land-use classification, plantation monitoring, and damage assessment of natural disasters. As the plantations can span vast areas, the manual identification of each tree is costly, tedious, and error-prone, and so automatic classification methods are necessary.

Classification of tree species in aerial images has been actively investigated~\cite{fassnacht2016review}. In \cite{puttemans2018comparing} and~\cite{aparna2018cnn}, the authors present automatic solutions to detect coconut trees based on convolutional neural network (CNN) models. Despite these recent advances, CNN models usually require considerable human effort in image annotation to create large training sets. Vargas-Muñoz et al.~\cite{8899005} propose an active learning approach to mitigate the problem. The method explores data projection techniques to allow simultaneous annotation of multiple patches as having or not coconut trees. It then uses a CNN to identify the candidate patches with coconut trees. We adopt another alternative -- the design of simplified CNN models from small training sets with user interaction, being the user an expert in the design of CNN models.

@@

In order to reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called FLIM (\emph{Feature Learning from Image Markers}), that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, a few training images, and draws markers in regions of the input images that best discriminate the classes. \textcolor{blue}{Classes may appear in distinct clusters of a local feature space defined by patches extracted from marker pixels at each convolutional layer's input.} FLIM defines the weights of the filters as the centroids of the clusters. Such filters are expected to activate discriminant regions in the given convolutional layer's output, and the process repeats in a layer-by-layer fashion. The architecture of the CNN could also be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}.

We demonstrate the advantages of this interactive technique on binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of marker selection.  Second, very few images per class (e.g., less than five) seem to be enough. \textcolor{red}{Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers.} Fourth, FLIM is application-independent, and so it might be useful in other image classification problems.

@@

\textcolor{blue}{A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation by eliminating negative activations, and aggregate the resulting activations locally (pooling) and globally (flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success.} Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and apply fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually relies on weight optimization using backpropagation, which requires a higher number of annotated images as deep as the model. In FLIM, we eliminate the need for backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect(P_I(p)) \cdot  \vect(F)$, where $P_I(p)$ is a patch with $k\times k$ pixels and $m$ image channels around $p$, $F$ is a filter with $k\times k\times m$ weights, and $\vect$ is the vectorization operation. A filter $F$ can discriminate a class $i\in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect(P_I(p))$ of class $i$ and negative results for patterns from other classes. That is, $\vect(F)$ is the normal vector of a hyperplane in $\Re^{k\times k\times m}$ which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect(P_I(p))$ from all the $c$ classes.  
\begin{eqnarray}
{\cal P} & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p)\in [1,c]} P_I(p)
\label{eq.pset}
\end{eqnarray}
The patch set ${\cal P}$ defines samples from all classes in $\Re^{k\times k\times m}$ and a clustering operation on ${\cal P}$ creates groups with similar local patterns. Ideally, each group should contain patterns of a same class. However, some classes might share similar patterns. We should at least expect that their patterns be detected in different image regions. The groups must be shifted to the origin of $\Re^{k\times k\times m}$ by subtracting the mean value of each feature in ${\cal P}$. Additionally, the standard deviation of each feature is computed and used for standardization. This operation allows batch normalization of image sets using the mean and standard deviation of the features from the image markers -- i.e.,  \emph{a marker-based batch normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$ and we force norm $\|\vect(F)\|=1$ to avoid preferences among them. The centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane with normal $\vect(F)$. Figure~\ref{fig:filter}a illustrates three groups with color-coded samples from two classes in a hypothetical 2D feature space. The marker-based batch normalization and convolution with filters obtained as centroids of the three groups creates a new 3D feature space (Figure~\ref{fig:filter}b), in which the classes can be more easily detected by two filters.

\begin{figure}[!ht]
  \centering
  \subfloat[\label{fig:ex-groups}]
  {\includegraphics[width=.35\linewidth]{figures/example_groups/groups_centered}}
  ~
  \subfloat[\label{fig:ex-groups-after}]{\includegraphics[width=.45\linewidth]{figures/example_groups/groups3d_centered}
  }
  \caption{(a) A 2D feature space with color-coded samples from two classes distributed in three groups. (b) After marker-based batch normalization and convolution with filters obtained from each group, the classes can be more easily detected by two filters in a new 3D feature space.}
  \label{fig:filter}
\end{figure}

Each convolutional layer is trained individually, one layer at a time, to find its filter set ${\cal F}$. The number of filters per layer depends on the clustering technique. We use $K$-means, then the user must specify the number $K=|{\cal F}|$ of filters. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max-pooling operation to aggregate local information. Note that, apart from the initial marker selection, the training process is automatic. We have preserved the convolutional layers' output dimensions and the marker pixels used to find the filters of layer $L+1$ from the output of layer $L$ are always the same for all convolutional layers. Figure~\ref{fig:tree} shows one example of a simple network projected by FLIM for the experiments of the next section.

@@

  \begin{tikzpicture}
    [
      grow                    = down,
      sibling distance        = 5em,
      level distance          = 3em,
      edge from parent/.style = {draw=black, -latex},
      every node/.style       = {treenode},
      sloped
    ]
    \node [treenode, fill=tealDeer] (v0) {MarkerBased BatchNorm}
    child {
        node [treenode, fill=tealDeer] {Conv$(60 \times 7 \times 7)$}
        child {
          node [treenode, fill=tealDeer] {ReLU} 
          child {
            node [treenode, fill=tealDeer] {MaxPool$(3\times3)$}
            child {
              node [treenode, fill=tealDeer] {BatchNorm}
              child {
                node [treenode, fill=navyPurple] {Classifier}
                edge from parent
              }
            }
          } 
        }
      };

@@

\section{Experiments}

We use a dataset from~\cite{8899005}, which contains aerial images with or without one coconut tree for counting. These images belong to a large imagery from

selected aerial images from WeRobotics and used the same dataset of regions of interest in these images, as provided by\cite{8899005} . The imagery with a resolution of 8cm was captured by the World Bank's UAVs for Disaster Resilience Program, collaborating with WeRobotics and OpenAerialMap in the Kingdom of Tonga in October 2017. The Humanitarian OpenStreetMap community provided the annotation. The dataset consists of 13587 patches of dimensions $90 \times 90$ of coconut (10268) and non-coconut (3319). The latter class has a large variability. The problem of interest is to classify the images in one of the two classes. We only report experiments with one dataset due to the space limitation.

@@

    FLIM & $0.863 \pm 0.002$ & $0.849 \pm 0.005$ & $0.854 \pm 0.004$\\
    FLIM-FT & $0.845 \pm 0.003$ & $0.825\pm 0.006$ & $0.832 \pm 0.005$ \\

    VGG & $0.827 \pm 0.003$ & $0.770 \pm 0.016$  & $ 0.784 \pm 0.014$\\
    VGG-FT & $0.872 \pm 0.007$ & $0.844 \pm 0.015$ & $0.851 \pm 0.014 $ \\
